Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-05 draft.tex
shortened 'Interactive Decision Tree Visualization' section

They present an interactive data mining tool, which can analyze association rules and highlight interesting relations. %, which may trigger new hypotheses about the data.

@@

@@

@@

These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.
%%
\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a established classification method in data mining to create predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
Readability is only granted for small trees.
%%
Large, complex trees are difficult to trace and to understand.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields the same amount of trees.
%%
Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
%%
We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
%%
This is included in our IVA framework.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
%%
A first analysis step was performed by applying the classification algorithm to the data.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error}.
%%
The two measures form the axes for a scatter plot of the classification results.
%%
The scatter plot is our central element for the interactive analysis of decision trees.
\\\\
\noindent \textbf{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.
\\\\
\noindent \textbf{Attribute Mapping.}
%%
The scatter plot axes are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.

Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
%%
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
%%
This makes these results less reliable.
%%
Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.
%%
The initial value is empirically set to $100$, marking a good tradeoff between sparse variables and statistical informative value.
%%
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
%%
This allows instant reliability assessment of the result.
%%

We apply a square root scale for the tree size axis to highlight decision trees with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
%%
Each dichotomous variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.
%%
This allows to sequentially analyze the classifications.

@@

\subsubsection{Implementation}
All analyses are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.

@@

As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
Furthermore, calculations based on a prior data selection can be redone within the report.

@@

Exchanging the prototype becomes as easy as exchanging a hyperlink.

@@

The prototype is available at \url{blind.dnsalias.com}.

@@

The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizationOfClassificationResults}).

>>>>

They present an interactive data mining tool, which can analyze association rules and highlight relations. %, which may trigger new hypotheses about the data.

@@

All categorical variables are converted into binary \emph{dummy variables}, indicate the presence or absence of a categorical variable manifestation. %with multiple manifestations into several dichotomous variables.
%Each dichotomous variable encodes the presence of a manifestation.
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%One subject can only have one of these variables set to true.
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.

@@

%% TODO: Shorten

@@

These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.
%%
% \section{\uppercase{Evaluation of Decision Trees}}
\section{\uppercase{Interactive Decision Tree Visualization}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a established classification method in data mining to create predictive models.
%%
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
%They work with numerical as well as categorical data.
%%
Readability is only granted for small trees.
%%
%Large, complex trees are difficult to trace and to understand.
%%
Too many branches impose overfitting to the data \cite{DecisionTree}.
%%
\\\\
\noindent \textbf{Creating Decision Trees.}
We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%Such a calculation requires a numeric or categorical target variable.
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.
\subsection{Visualizing Decision Trees}
\label{subsec:VisualizingDecisionTrees}
We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
Since we have 134 non-image variables, the calculation yields the same amount of trees.
Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
This is included in our IVA framework.
\noindent \textbf{Visualization of Classification Results.}
We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
A first analysis step was performed by applying the classification algorithm to the data.
The optimal classification uses a few rules to precisely characterize the target variable.
Therefore, we are interested in \emph{small trees} with a \emph{low classification error}.
The two measures form the axes for a scatter plot of the classification results.
The scatter plot is our central element for the interactive analysis of decision trees.
\noindent \textbf{The Error Term.}
Calculating the mean classification error is imprecise for non-uniform distributions.
%Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%The metric usually works well for variables with uniform distribution.
%It distorts the result for other distribution types.
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, we get an mean error of 10\%, even though it is very bad.
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
\begin{equation}
totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
\end{equation}
$M$ represents the set of manifestation of each variable.
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
It allows for comparability of error rates between variables with different manifestation count.
\textbf{Attribute Mapping.}
The scatter plot axes are defined by \emph{tree size} and the previously described \emph{error metric}.
This allows us to visualize a multitude of classification results in one plot.
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.
Many interdependent variables are sparse, such as \emph{medication of diabetes} or \emph{reason of early retirement}.
The classification algorithm may produce higher accuracy for variables with less subjects due to the small sample size.
This makes these results less reliable.
Therefore, we provide a way to adjust the minimal number of subjects for each variable using a slider input.
The initial value is empirically set to $100$, marking a good tradeoff between sparse variables and statistical informative value.
Furthermore, we map the number of subjects associated with a variable to point diameter in the scatter plot.
This allows instant reliability assessment of the result.
We apply a square root scale for the tree size axis to highlight decision trees with few decision rules.
Outlier results with very large decision trees would otherwise distort the resulting plot.
\subsection{Interaction With the Decision Tree Visualization}
The described visualization provides a good overview over the classification results.
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.
This allows to sequentially analyze the classifications.

@@

\noindent \textbf{Implementation.}
All analyses are available at \url{blind.dnsalias.com} and are carried out using \texttt{R}, a widely used programming language for statistical calculations and visualizations.

@@

@@

% As opposed to the standard \texttt{R} plots, \texttt{ggvis} allows to adjust visualization variables using user interface controls, such as sliders.
% In order to make the train of thought comprehensible, we used \texttt{RMarkdown}, which allows to create reports by combining \texttt{R} with the \texttt{Markdown} syntax.
% We used \texttt{R Shiny}\footnote{Developed by RStudio, Inc; \texttt{shiny.rstudio.com}} to make the report available as dynamic web application.
% It allows to combine the power of static \texttt{R Markdown} reports with dynamically parameterized \texttt{ggvis} plots.
% Furthermore, calculations based on a prior data selection can be redone within the report.

@@

% The web-based approach allows us to quickly exchange results with our collaborating epidemiological experts.
% They can use the technique without installing any software.
% Exchanging the prototype becomes as easy as exchanging a hyperlink.
% The prototype is available at \url{blind.dnsalias.com}.

@@

The x-axis shows the number of decisions of the underlying model, the y-axis the classification error (see Section~\ref{subsec:VisualizingDecisionTrees}).
